Type Classes for Mathematics in Type Theory

### 2010-11-23 papers/mscs/mscs.tex
Replace Π's with ∀'s in MSCS paper.

% \documentclass{mscs}
\documentclass[a4paper,10pt,runningheads]{llncs}
\usepackage{url}
\usepackage{bbm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{upgreek}
\usepackage[mathletters]{ucs}
\usepackage[utf8x]{inputenx}
\usepackage{comment}
\newcommand{\N}{\ensuremath{\mathbbm{N}}}
\newcommand{\Z}{\ensuremath{\mathbbm{Z}}}
\newcommand{\Q}{\ensuremath{\mathbbm{Q}}}
\usepackage{color}

@@

\def\lstlanguagefiles{defManSSR.tex}
\lstset{language=SSR}
\lstset{literate=
  % symbols that actually occur as unicode in our source:
  {λ}{{$\uplambda\ $}}1
  {σ}{{$\upsigma$}}1
  {η}{{$\upeta$}}1
  {φ}{{$\upphi$}}1
  {∃}{{$\exists$}}1
  {→}{{$\to\ $}}1
  {≠}{{$\ne\ $}}1
  {¬}{{$\neg\ $}}1
  {⟶}{{$\longrightarrow\ $}}1
  {⇛}{{$\Longrightarrow\ $}}1
  {∧}{{$\land$}}1
  {∀}{{$\forall$}}1
  {Π}{{$\Uppi$ }}1
  {η}{{$\upeta$}}1
  {⊓}{{$\sqcap$}}1
  {∘}{{$\circ\ $}}1
  {◎}{{$\odot$}}1
  { ≡ }{{$\equiv\ $}}1
  % things we can't make pretty in the actual source, but can make pretty here!:
  {=>}{{$\,\Rightarrow\ $}}1
  {==>}{{$\Rightarrow\ $}}1
  {<-}{{$\leftarrow\ $}}1
  {~}{{$\sim$}}1
}

@@

The `native' notion of equality in Coq is that of terms being convertible, naturally reified as a proposition by the inductive type family \lstinline|eq| with single constructor \lstinline|eq_refl|:
\begin{lstlisting}
  eq_refl : Π (T: Type) (x: T), x ≡ x,
\end{lstlisting}
where `\lstinline|a ≡ b|' is notation for \lstinline|eq T a b|. Here we diverge from Coq tradition and reserve \lstinline|=| for setoid equality, as this is the equality we will be working with most of the time. Importantly, since convertibility is a congruence, a proof of \lstinline|a ≡ b| lets us substitute \lstinline|b| for \lstinline|a| anywhere inside a term without further conditions. We mention this explicitly only because rewriting \emph{does} give rise to conditions when we depart from Leibniz equality and introduce equivalence relations representing abstractions over common representations of the same conceptual object. For instance, consider formal fractions of integers as a representation of rationals. Rewriting a subterm using such an equality is permitted only if the subterm is an argument of a function that has been proven to \emph{respect} the equality. Such a function is called \emph{proper}, and propriety must be proved for each function in whose arguments we wish to enable rewriting.

@@

\begin{lstlisting}
  Definition reflexive {A: Type} (R: relation A): Prop := Π a, R a a.
\end{lstlisting}
The curly brackets used for \lstinline|A| mark it as an implicit argument.

@@

In the absence of type classes, these are all very real problems, and for this reason the two largest formalizations of abstract algebraic structures in Coq today, CoRN~\cite{C-corn} and \textsc{Ssreflect}~\cite{Packed}, both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters. For reflexive relations, the following is a fully bundled representation---the other end of the spectrum:
\begin{lstlisting}
  Record ReflexiveRelation: Type :=
    { rr_carrier: Type
    ; rr_rel: relation rr_carrier
    ; rr_proof: Π x, rr_rel x x }.
\end{lstlisting}
Superficially, this instantly solves the problems described above: reflexive relations can now be declared and passed as self-contained packages, and the \lstinline|rr_rel| projection now constitutes a canonical name for relations that are known to be reflexive, and we could bind a notation to it. While there is no conventional notation for reflexive relations, the situation is the same in the context of, say, a semiring, where we would bind $+$ and $*$ notations to the record field projections for the addition and multiplication operations, respectively.

@@

Without further support from the system, this would have to be written as
\begin{lstlisting}
  Theorem example G e op {P: SemiGroup G e op}:
    Π x y z, e (op (op x y) z) (op (op z y) x).
\end{lstlisting}
Because \lstinline|e| and \lstinline{op} are freshly introduced local names, we cannot bind notations to them prior to this theorem. Hence, if we want notations, what we really need are canonical names for these components. This is easily accomplished with single-field type classes containing one component each, which we will call \emph{operational type classes}\footnote{These single-field type classes are used in the same way in the \lstinline|Clean| standard library~\cite{clean}.}:
\begin{lstlisting}
  Class Equiv A := equiv: relation A.
  Class SemiGroupOp A := sg_op: A $\to$ A $\to$ A.

@@

If we now retype \lstinline|SemiGroup| as:
\begin{lstlisting}
  Π (G: Type) (e: Equiv G) (op: SemiGroupOp G), Prop
\end{lstlisting}
then we can declare the theorem with:
\begin{lstlisting}
  Theorem example G e op {P: SemiGroup G e op}:
    Π x y z, x & y & z = z & y & x.
\end{lstlisting}
This works because instance resolution, invoked by the use of \lstinline|=| and \lstinline|&|, will find \lstinline|e| and \lstinline|op|, respectively. Hence, the above is really

\begin{lstlisting}
  Theorem example G e op {P: SemiGroup G e op}:
    Π x y z, equiv e (sg_op op (sg_op op x y) z) (sg_op op (sg_op op z y) x).
\end{lstlisting}
where \lstinline|e| and the \lstinline|op|s are filled in by instance resolution.

@@

A remaining eyesore in the theorem declaration is the enumeration of \lstinline|e| and \lstinline|op|. To remove these, we use a new parameter declaration feature called \emph{implicit generalization}, introduced in Coq specifically to support type classes. Using implicit generalization, we can write:
\begin{lstlisting}
  Theorem example `{SemiGroup G}: Π x y z: G, x & y & z = z & y & x.
\end{lstlisting}
The backtick tells Coq to insert implicit declarations of further parameters to \lstinline|SemiGroup G|, namely those declared as \lstinline|e| and \lstinline|op| above. It further lets us omit a name for the \lstinline|SemiGroup G| parameter itself as well. All of these will be given automatically generated names that we will never refer to.

@@

\lstinline|Decision| is a very general-purpose type class that also works for predicates. For instance, to declare a parameter expressing decidability of, say, (setoid) equality on a type \lstinline|X|, we write: \lstinline|`{Π a b: X, Decision (a = b)}|. To then use this (unnamed) decider to decide a particular equality, we simply say \lstinline|decide (x = y)|, and instance resolution will resolve the decider we declared.

@@

Following our idiom, we introduce operational type classes for the \emph{components} of a category:
\begin{lstlisting}
  Class Arrows (O: Type): Type := Arrow: O → O → Type.
  Class CatId O `{Arrows O} := cat_id: `(x ⟶ x).
  Class CatComp O `{Arrows O} :=
    comp: Π {x y z}, (y ⟶ z) → (x ⟶ y) → (x ⟶ z).

@@

With these in place, our type class for categories follows the usual type-theoretical definition of a
category~\cite{saibi1995constructive}:
\begin{lstlisting}
  Class Category (O: Type) `{Arrows O} `{Π x y: O, Equiv (x ⟶ y)}
      `{CatId O} `{CatComp O}: Prop :=
    { arrow_equiv:> Π x y, Setoid (x ⟶ y)
    ; comp_proper:> Π x y z, Proper (equiv ==> equiv ==> equiv) comp
    ; comp_assoc w x y z (a: w ⟶ x) (b: x ⟶ y) (c: y ⟶ z):
        c ◎$$ (b ◎$$ a) = (c ◎$$ b) ◎$$ a
    ; id_l `(a: x ⟶ y): cat_id ◎$$ a = a
    ; id_r `(a: x ⟶ y): a ◎$$ cat_id = a }.
\end{lstlisting}
This definition is based on the 2-categorical idea of having equality only on arrows, not on objects.
%Similarly, we will have equality on natural transformations, but not on functors. % todo: ORLY?

Initiality, too, is defined by a combination of an operational and a predicate class:
\begin{lstlisting}
  Context `{Category X}.
  Class InitialArrows (x: X): Type := initial_arrow: Π y, x ⟶ y.
  Class Initial (x: X) `{InitialArrows x}: Prop :=
    initial_arrow_unique: Π y (a: x ⟶ y), a = initial_arrow y.
\end{lstlisting}
The operational class \lstinline|InitialArrows| designates the arrows that originate from an initial object \lstinline|x| by virtue of it being initial. The \lstinline|Initial| class itself further requires these ``initial arrows'' to be unique. Having \lstinline|InitialArrows| as an operational type class means that we can always simply say \lstinline|initial_arrow y| whenever \lstinline|y| is known to be an object in a category known to have an initial object (where `known' should be read as ``can be determined by instance resolution'').

@@

  Class Fmap: Type :=
    fmap: Π {v w: C}, (v ⟶ w) → (map_obj v ⟶ map_obj w).

  Class Functor `{Fmap}: Prop :=
    { functor_from: Category C
    ; functor_to: Category D
    ; functor_morphism:> Π a b: C, Setoid_Morphism (@fmap _ a b)
    ; preserves_id: `(fmap (cat_id: a ⟶ a) = cat_id)
    ; preserves_comp `(f: y ⟶ z) `(g: x ⟶ y):
      fmap (f ◎$$ g) = fmap f ◎$$ fmap g }.
\end{lstlisting}
We ought to say a few words about our use of \lstinline|fmap|. The usual mathematical notational convention for functor application is to use the name of the functor to refer to both its object map and its arrow map, relying on additional conventions regarding object/arrow names for disambiguation: \lstinline|F x| and \lstinline|F f| map an object and an arrow, respectively, because \lstinline|x| and \lstinline|f| are conventional names for objects and arrows, respectively.

In Coq, for a term \lstinline|F| to function as though it had two different types simultaneously (namely the object map and the arrow map), there must either (1) be coercions from the type of F to either function, or (2) F must be (coercible to) a single function that is able to consume both object and arrow arguments. In addition to not being supported by Coq, option (1) would violate our policy of leaving components unbundled. For (2), if it could be made to work at all, F would need a pretty egregious type considering that arrow types are indexed by objects, and that the type of the arrow map
\begin{lstlisting}
 Π x y, (x ⟶ y) → (F x ⟶ F y)
\end{lstlisting}
must refer to the object map.

@@

We introduce a convenient notation for the type of the computational content of a natural transformation between two functors:
\begin{lstlisting}
  Notation "F ⇛ G" := (Π x, F x ⟶ G x).
\end{lstlisting}
Now assume the following context:
\begin{lstlisting}
  Context `{Category C} `{Category D}
    `{Functor (F: C → D)} `{Functor (G: D → C)}.
\end{lstlisting}
The naturality property is easy to write:
\begin{lstlisting}
  Class NaturalTransformation (η: F ⇛ G): Prop :=
    { naturaltrans_from: Functor F
    ; naturaltrans_to: Functor G
    ; natural: Π `(f: x ⟶ y), η$$ y ◎$$ fmap F f = fmap G f ◎$$ η$$ x }.
\end{lstlisting}
  % Todo: Having to add "$$" to get whitespace is awful. we can't add "\ " in η's literate-replacement though, because then we also get space before the colon in "η: t".

@@

Adjunctions can be defined in different ways. A nice symmetric definition is the following:
\begin{lstlisting}
  Class Adjunction (φ: Π `(F c ⟶ d), (c ⟶ G d)): Prop :=
    { adjunction_left_functor: Functor F _
    ; adjunction_right_functor: Functor G _
    ; natural_left `(f: d ⟶ d') c: (fmap G f ◎) ∘ φ$$ = φ (c:=c) ∘ (f ◎)
    ; natural_right `(f: c' ⟶ c) d: (◎ f) ∘ φ (d:=d) = φ$$ ∘ (◎ fmap F f) }.
\end{lstlisting}
An alternative definition is the following:
\begin{lstlisting}
  Class AltAdjunction (η: id ⇛ G ∘ F) (φ: Π `(f: c ⟶ G d), F c ⟶ d): Prop :=
    { alt_adjunction_natural_unit: NaturalTransformation η
    ; alt_adjunction_factor: Π `(f: c ⟶ G d),
        is_sole ((f =) ∘ (◎ η$$ c) ∘ fmap G) (φ f) }.
\end{lstlisting}
Formalizing the (nontrivial) proof that these two definitions are equivalent provides a nice test for our definitions. As a first step, we have constructed the unit and co-unit of the adjunction, thus proving MacLane's Theorem 1. We have concisely and closely followed his proof~\cite{CatWork}. % todo: ref?

@@

  Class AlgebraOps (σ: Signature) (carriers: sorts σ$$ → Type)
    := algebra_op: Π o: operation σ, op_type carriers (σ o).
\end{lstlisting}
Because our carriers will normally be equipped with a setoid equality, we further define the predicate class \lstinline|Algebra|, stating that each of the operations respects the setoid equality on the carriers:
\begin{lstlisting}
  Class Algebra `{Π a, Equiv (carriers a)} `{AlgebraOps σ$$ carriers}: Prop :=
      { algebra_setoids:> Π a, Setoid (carriers a)
      ; algebra_propers:> Π o: σ, Proper (=) (algebra_op o) }.
\end{lstlisting}
The \lstinline|(=)| referred to in \lstinline|algebra_propers| is an \lstinline|Equiv| instance expressing setoid-respecting extensionality for the function types generated by \lstinline|op_type|.

@@

We now introduce a predicate class designating algebras that satisfy the laws of an equational theory:
\begin{lstlisting}
  Class InVariety
    (et: EquationalTheory) (carriers: sorts et → Type)
    {e: Π a, Equiv (carriers a)} `{AlgebraOps et carriers}: Prop :=
    { variety_algebra:> Algebra et carriers
    ; variety_laws: Π s, eqt_laws et s → (Π $\hspace{-1.5mm}$vars, eval_stmt et vars s) }.
\end{lstlisting}
  % todo: why do i need the manual \hspace!?

@@

  Record ObjectInVariety: Type := object_in_variety
    { variety_carriers:> sorts et → Type
    ; variety_equiv: Π a, Equiv (variety_carriers a)
    ; variety_op: AlgebraOps et variety_carriers
    ; variety_proof: InVariety et variety_carriers }.
\end{lstlisting}
The arrows will be homomorphisms, which are also defined generically for any equational theory:

@@

A set of relations \lstinline|e| (one for each sort) is a congruence for an existing algebra if (1) \lstinline|e| respects that algebra's existing setoid equality, and (2) the operations with \lstinline|e| again form an algebra (namely the quotient algebra):
\begin{lstlisting}
  Context `{Algebra σ$$ A}.
  Class Congruence (e: Π s: sorts σ, relation (v s)): Prop :=
    { congruence_proper:> Π s, Proper (equiv ==> equiv ==> iff) (e s)
    ; congruence_quotient:> Algebra σ$$ v (e:=e) }.
\end{lstlisting}
We have proved that this natural and economical type-theoretic formulation that leverages our systematic integration of setoid equality is equivalent to the traditional definition of congruences as relations that, represented as sets of pairs, form a subalgebra of the product algebra.

@@

  Instance: ClosedSubset image.
\end{lstlisting}
The \lstinline|sigT| type constructor is a \lstinline|Type|-sorted existential quantifier. \lstinline|ClosedSubset| is defined elsewhere as:
\begin{lstlisting}
  Context `{Algebra σ$$ A} (P: Π s, A s → Type).
  Class ClosedSubset: Type :=
    { subset_proper: Π s x x', x = x' → iffT (P s x) (P s x')
    ; subset_closed: Π o, op_closed (algebra_op o) }.
\end{lstlisting}
Here, \lstinline|op_closed| is defined by recursion over the symbolic operation types.

@@

Taking these conversion functions for granted, we will also need a ``down to earth'' representation of the initiality arrows if we are to give a \lstinline|SemiRing|-based definition of the interface for natural numbers. Once again, we introduce an operational type class to represent this particular component:
\begin{lstlisting}
  Class NaturalsToSemiRing (A: Type) :=
    naturals_to_semiring: Π B `{RingMult B} `{RingPlus B} `{RingOne B}
      `{RingZero B}, A → B.
\end{lstlisting}
The instance for \lstinline|nat| is defined as follows:
\begin{lstlisting}
Instance nat_to_semiring: NaturalsToSemiRing nat :=
  λ _ _ _ _ _ => fix f (n: nat) := match n with 0 => 0 | S m => f m + 1 end.
\end{lstlisting}

@@

With these instances in place, we can now define the improved natural numbers specification:
\begin{lstlisting}
  Context `{SemiRing A} `{NaturalsToSemiRing A}.
  Class Naturals: Prop :=
    { naturals_ring:> SemiRing A
    ; naturals_to_semiring_mor:> Π `{SemiRing B},
        SemiRing_Morphism (naturals_to_semiring A B)
    ; naturals_initial:> Initial (bundle_semiring A) }.
\end{lstlisting}
Basing theory and programs on this abstract interface instead of on specific implementation (such as the ubiquitous Peano naturals \lstinline|nat| in the Coq standard library) is not only cleaner mathematically, but also facilitates easy swapping between implementations. And this benefit is far from theoretical, as diverse representations of the natural numbers are abound; for instance, unary, binary, factor multisets, and arrays of native machine words.

>>>>

% \documentclass{mscs}
\documentclass[a4paper,10pt,runningheads]{llncs}
\usepackage{url}
\usepackage{bbm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{upgreek}
\usepackage[mathletters]{ucs}
\usepackage[utf8x]{inputenx}
\usepackage{comment}
\newcommand{\N}{\ensuremath{\mathbbm{N}}}
\newcommand{\Z}{\ensuremath{\mathbbm{Z}}}
\newcommand{\Q}{\ensuremath{\mathbbm{Q}}}

\usepackage{color}

@@

\def\lstlanguagefiles{defManSSR.tex}
\lstset{language=SSR}
\lstset{literate=
  % symbols that actually occur as unicode in our source:
  {λ}{{$\uplambda\ $}}1
  {σ}{{$\upsigma$}}1
  {η}{{$\upeta$}}1
  {φ}{{$\upphi$}}1
  {∃}{{$\exists$}}1
  {→}{{$\to\ $}}1
  {≠}{{$\ne\ $}}1
  {¬}{{$\neg\ $}}1
  {⟶}{{$\longrightarrow\ $}}1
  {⇛}{{$\Longrightarrow\ $}}1
  {∧}{{$\land$}}1
  {∀}{{$\forall\ $}}1
  {η}{{$\upeta$}}1
  {⊓}{{$\sqcap$}}1
  {∘}{{$\circ\ $}}1
  {◎}{{$\odot$}}1
  { ≡ }{{$\equiv\ $}}1
  % things we can't make pretty in the actual source, but can make pretty here!:
  {=>}{{$\,\Rightarrow\ $}}1
  {==>}{{$\Rightarrow\ $}}1
  {<-}{{$\leftarrow\ $}}1
  {~}{{$\sim$}}1
}

@@

The `native' notion of equality in Coq is that of terms being convertible, naturally reified as a proposition by the inductive type family \lstinline|eq| with single constructor \lstinline|eq_refl|:
\begin{lstlisting}
  eq_refl : ∀ (T: Type) (x: T), x ≡ x,
\end{lstlisting}
where `\lstinline|a ≡ b|' is notation for \lstinline|eq T a b|. Here we diverge from Coq tradition and reserve \lstinline|=| for setoid equality, as this is the equality we will be working with most of the time. Importantly, since convertibility is a congruence, a proof of \lstinline|a ≡ b| lets us substitute \lstinline|b| for \lstinline|a| anywhere inside a term without further conditions. We mention this explicitly only because rewriting \emph{does} give rise to conditions when we depart from Leibniz equality and introduce equivalence relations representing abstractions over common representations of the same conceptual object. For instance, consider formal fractions of integers as a representation of rationals. Rewriting a subterm using such an equality is permitted only if the subterm is an argument of a function that has been proven to \emph{respect} the equality. Such a function is called \emph{proper}, and propriety must be proved for each function in whose arguments we wish to enable rewriting.

@@

\begin{lstlisting}
  Definition reflexive {A: Type} (R: relation A): Prop := ∀ a, R a a.
\end{lstlisting}
The curly brackets used for \lstinline|A| mark it as an implicit argument.

@@

In the absence of type classes, these are all very real problems, and for this reason the two largest formalizations of abstract algebraic structures in Coq today, CoRN~\cite{C-corn} and \textsc{Ssreflect}~\cite{Packed}, both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters. For reflexive relations, the following is a fully bundled representation---the other end of the spectrum:
\begin{lstlisting}
  Record ReflexiveRelation: Type :=
    { rr_carrier: Type
    ; rr_rel: relation rr_carrier
    ; rr_proof: ∀ x, rr_rel x x }.
\end{lstlisting}
Superficially, this instantly solves the problems described above: reflexive relations can now be declared and passed as self-contained packages, and the \lstinline|rr_rel| projection now constitutes a canonical name for relations that are known to be reflexive, and we could bind a notation to it. While there is no conventional notation for reflexive relations, the situation is the same in the context of, say, a semiring, where we would bind $+$ and $*$ notations to the record field projections for the addition and multiplication operations, respectively.

@@

Without further support from the system, this would have to be written as
\begin{lstlisting}
  Theorem example G e op {P: SemiGroup G e op}:
    ∀ x y z, e (op (op x y) z) (op (op z y) x).
\end{lstlisting}
Because \lstinline|e| and \lstinline{op} are freshly introduced local names, we cannot bind notations to them prior to this theorem. Hence, if we want notations, what we really need are canonical names for these components. This is easily accomplished with single-field type classes containing one component each, which we will call \emph{operational type classes}\footnote{These single-field type classes are used in the same way in the \lstinline|Clean| standard library~\cite{clean}.}:
\begin{lstlisting}
  Class Equiv A := equiv: relation A.
  Class SemiGroupOp A := sg_op: A $\to$ A $\to$ A.

@@

If we now retype \lstinline|SemiGroup| as:
\begin{lstlisting}
  ∀ (G: Type) (e: Equiv G) (op: SemiGroupOp G), Prop
\end{lstlisting}
then we can declare the theorem with:
\begin{lstlisting}
  Theorem example G e op {P: SemiGroup G e op}:
    ∀ x y z, x & y & z = z & y & x.
\end{lstlisting}
This works because instance resolution, invoked by the use of \lstinline|=| and \lstinline|&|, will find \lstinline|e| and \lstinline|op|, respectively. Hence, the above is really

\begin{lstlisting}
  Theorem example G e op {P: SemiGroup G e op}:
    ∀ x y z, equiv e (sg_op op (sg_op op x y) z) (sg_op op (sg_op op z y) x).
\end{lstlisting}
where \lstinline|e| and the \lstinline|op|s are filled in by instance resolution.

@@

A remaining eyesore in the theorem declaration is the enumeration of \lstinline|e| and \lstinline|op|. To remove these, we use a new parameter declaration feature called \emph{implicit generalization}, introduced in Coq specifically to support type classes. Using implicit generalization, we can write:
\begin{lstlisting}
  Theorem example `{SemiGroup G}: ∀ x y z: G, x & y & z = z & y & x.
\end{lstlisting}
The backtick tells Coq to insert implicit declarations of further parameters to \lstinline|SemiGroup G|, namely those declared as \lstinline|e| and \lstinline|op| above. It further lets us omit a name for the \lstinline|SemiGroup G| parameter itself as well. All of these will be given automatically generated names that we will never refer to.

@@

\lstinline|Decision| is a very general-purpose type class that also works for predicates. For instance, to declare a parameter expressing decidability of, say, (setoid) equality on a type \lstinline|X|, we write: \lstinline|`{∀ a b: X, Decision (a = b)}|. To then use this (unnamed) decider to decide a particular equality, we simply say \lstinline|decide (x = y)|, and instance resolution will resolve the decider we declared.

@@

Following our idiom, we introduce operational type classes for the \emph{components} of a category:
\begin{lstlisting}
  Class Arrows (O: Type): Type := Arrow: O → O → Type.
  Class CatId O `{Arrows O} := cat_id: `(x ⟶ x).
  Class CatComp O `{Arrows O} :=
    comp: ∀ {x y z}, (y ⟶ z) → (x ⟶ y) → (x ⟶ z).

@@

With these in place, our type class for categories follows the usual type-theoretical definition of a
category~\cite{saibi1995constructive}:
\begin{lstlisting}
  Class Category (O: Type) `{Arrows O} `{∀$$x y: O, Equiv (x ⟶ y)}
      `{CatId O} `{CatComp O}: Prop :=
    { arrow_equiv:> ∀ x y, Setoid (x ⟶ y)
    ; comp_proper:> ∀ x y z, Proper (equiv ==> equiv ==> equiv) comp
    ; comp_assoc w x y z (a: w ⟶ x) (b: x ⟶ y) (c: y ⟶ z):
        c ◎$$ (b ◎$$ a) = (c ◎$$ b) ◎$$ a
    ; id_l `(a: x ⟶ y): cat_id ◎$$ a = a
    ; id_r `(a: x ⟶ y): a ◎$$ cat_id = a }.
\end{lstlisting}
This definition is based on the 2-categorical idea of having equality only on arrows, not on objects.
%Similarly, we will have equality on natural transformations, but not on functors. % todo: ORLY?

Initiality, too, is defined by a combination of an operational and a predicate class:
\begin{lstlisting}
  Context `{Category X}.
  Class InitialArrows (x: X): Type := initial_arrow: ∀ y, x ⟶ y.
  Class Initial (x: X) `{InitialArrows x}: Prop :=
    initial_arrow_unique: ∀ y (a: x ⟶ y), a = initial_arrow y.
\end{lstlisting}
The operational class \lstinline|InitialArrows| designates the arrows that originate from an initial object \lstinline|x| by virtue of it being initial. The \lstinline|Initial| class itself further requires these ``initial arrows'' to be unique. Having \lstinline|InitialArrows| as an operational type class means that we can always simply say \lstinline|initial_arrow y| whenever \lstinline|y| is known to be an object in a category known to have an initial object (where `known' should be read as ``can be determined by instance resolution'').

@@

  Class Fmap: Type :=
    fmap: ∀ {v w: C}, (v ⟶ w) → (map_obj v ⟶ map_obj w).

  Class Functor `{Fmap}: Prop :=
    { functor_from: Category C
    ; functor_to: Category D
    ; functor_morphism:> ∀ a b: C, Setoid_Morphism (@fmap _ a b)
    ; preserves_id: `(fmap (cat_id: a ⟶ a) = cat_id)
    ; preserves_comp `(f: y ⟶ z) `(g: x ⟶ y):
      fmap (f ◎$$ g) = fmap f ◎$$ fmap g }.
\end{lstlisting}
We ought to say a few words about our use of \lstinline|fmap|. The usual mathematical notational convention for functor application is to use the name of the functor to refer to both its object map and its arrow map, relying on additional conventions regarding object/arrow names for disambiguation: \lstinline|F x| and \lstinline|F f| map an object and an arrow, respectively, because \lstinline|x| and \lstinline|f| are conventional names for objects and arrows, respectively.

In Coq, for a term \lstinline|F| to function as though it had two different types simultaneously (namely the object map and the arrow map), there must either (1) be coercions from the type of F to either function, or (2) F must be (coercible to) a single function that is able to consume both object and arrow arguments. In addition to not being supported by Coq, option (1) would violate our policy of leaving components unbundled. For (2), if it could be made to work at all, F would need a pretty egregious type considering that arrow types are indexed by objects, and that the type of the arrow map
\begin{lstlisting}
 ∀ x y, (x ⟶ y) → (F x ⟶ F y)
\end{lstlisting}
must refer to the object map.

@@

We introduce a convenient notation for the type of the computational content of a natural transformation between two functors:
\begin{lstlisting}
  Notation "F ⇛ G" := (∀ x, F x ⟶ G x).
\end{lstlisting}
Now assume the following context:
\begin{lstlisting}
  Context `{Category C} `{Category D}
    `{Functor (F: C → D)} `{Functor (G: D → C)}.
\end{lstlisting}
The naturality property is easy to write:
\begin{lstlisting}
  Class NaturalTransformation (η: F ⇛ G): Prop :=
    { naturaltrans_from: Functor F
    ; naturaltrans_to: Functor G
    ; natural: ∀ `(f: x ⟶ y), η$$ y ◎$$ fmap F f = fmap G f ◎$$ η$$ x }.
\end{lstlisting}
  % Todo: Having to add "$$" to get whitespace is awful. we can't add "\ " in η's literate-replacement though, because then we also get space before the colon in "η: t".

@@

Adjunctions can be defined in different ways. A nice symmetric definition is the following:
\begin{lstlisting}
  Class Adjunction (φ: ∀ `(F c ⟶ d), (c ⟶ G d)): Prop :=
    { adjunction_left_functor: Functor F _
    ; adjunction_right_functor: Functor G _
    ; natural_left `(f: d ⟶ d') c: (fmap G f ◎) ∘ φ$$ = φ (c:=c) ∘ (f ◎)
    ; natural_right `(f: c' ⟶ c) d: (◎ f) ∘ φ (d:=d) = φ$$ ∘ (◎ fmap F f) }.
\end{lstlisting}
An alternative definition is the following:
\begin{lstlisting}
  Class AltAdjunction (η: id ⇛ G ∘ F) (φ: ∀ `(f: c ⟶ G d), F c ⟶ d): Prop :=
    { alt_adjunction_natural_unit: NaturalTransformation η
    ; alt_adjunction_factor: ∀ `(f: c ⟶ G d),
        is_sole ((f =) ∘ (◎ η$$ c) ∘ fmap G) (φ f) }.
\end{lstlisting}
Formalizing the (nontrivial) proof that these two definitions are equivalent provides a nice test for our definitions. As a first step, we have constructed the unit and co-unit of the adjunction, thus proving MacLane's Theorem 1. We have concisely and closely followed his proof~\cite{CatWork}. % todo: ref?

@@

  Class AlgebraOps (σ: Signature) (carriers: sorts σ$$ → Type)
    := algebra_op: ∀ o: operation σ, op_type carriers (σ o).
\end{lstlisting}
Because our carriers will normally be equipped with a setoid equality, we further define the predicate class \lstinline|Algebra|, stating that each of the operations respects the setoid equality on the carriers:
\begin{lstlisting}
  Class Algebra `{∀$$a, Equiv (carriers a)} `{AlgebraOps σ$$ carriers}: Prop :=
      { algebra_setoids:> ∀ a, Setoid (carriers a)
      ; algebra_propers:> ∀ o: σ, Proper (=) (algebra_op o) }.
\end{lstlisting}
The \lstinline|(=)| referred to in \lstinline|algebra_propers| is an \lstinline|Equiv| instance expressing setoid-respecting extensionality for the function types generated by \lstinline|op_type|.

@@

We now introduce a predicate class designating algebras that satisfy the laws of an equational theory:
\begin{lstlisting}
  Class InVariety
    (et: EquationalTheory) (carriers: sorts et → Type)
    {e: ∀ a, Equiv (carriers a)} `{AlgebraOps et carriers}: Prop :=
    { variety_algebra:> Algebra et carriers
    ; variety_laws: ∀ s, eqt_laws et s → (∀$$vars, eval_stmt et vars s) }.
\end{lstlisting}

@@

  Record ObjectInVariety: Type := object_in_variety
    { variety_carriers:> sorts et → Type
    ; variety_equiv: ∀ a, Equiv (variety_carriers a)
    ; variety_op: AlgebraOps et variety_carriers
    ; variety_proof: InVariety et variety_carriers }.
\end{lstlisting}
The arrows will be homomorphisms, which are also defined generically for any equational theory:

@@

A set of relations \lstinline|e| (one for each sort) is a congruence for an existing algebra if (1) \lstinline|e| respects that algebra's existing setoid equality, and (2) the operations with \lstinline|e| again form an algebra (namely the quotient algebra):
\begin{lstlisting}
  Context `{Algebra σ$$ A}.
  Class Congruence (e: ∀ s: sorts σ, relation (v s)): Prop :=
    { congruence_proper:> ∀ s, Proper (equiv ==> equiv ==> iff) (e s)
    ; congruence_quotient:> Algebra σ$$ v (e:=e) }.
\end{lstlisting}
We have proved that this natural and economical type-theoretic formulation that leverages our systematic integration of setoid equality is equivalent to the traditional definition of congruences as relations that, represented as sets of pairs, form a subalgebra of the product algebra.

@@

  Instance: ClosedSubset image.
\end{lstlisting}
The \lstinline|sigT| type constructor is a \lstinline|Type|-sorted existential quantifier. \lstinline|ClosedSubset| is defined elsewhere as:
\begin{lstlisting}
  Context `{Algebra σ$$ A} (P: ∀ s, A s → Type).
  Class ClosedSubset: Type :=
    { subset_proper: ∀ s x x', x = x' → iffT (P s x) (P s x')
    ; subset_closed: ∀ o, op_closed (algebra_op o) }.
\end{lstlisting}
Here, \lstinline|op_closed| is defined by recursion over the symbolic operation types.

@@

Taking these conversion functions for granted, we will also need a ``down to earth'' representation of the initiality arrows if we are to give a \lstinline|SemiRing|-based definition of the interface for natural numbers. Once again, we introduce an operational type class to represent this particular component:
\begin{lstlisting}
  Class NaturalsToSemiRing (A: Type) :=
    naturals_to_semiring: ∀ B `{RingMult B} `{RingPlus B} `{RingOne B}
      `{RingZero B}, A → B.
\end{lstlisting}
The instance for \lstinline|nat| is defined as follows:
\begin{lstlisting}
Instance nat_to_semiring: NaturalsToSemiRing nat :=
  λ _ _ _ _ _ => fix f (n: nat) := match n with 0 => 0 | S m => f m + 1 end.
\end{lstlisting}

@@

With these instances in place, we can now define the improved natural numbers specification:
\begin{lstlisting}
  Context `{SemiRing A} `{NaturalsToSemiRing A}.
  Class Naturals: Prop :=
    { naturals_ring:> SemiRing A
    ; naturals_to_semiring_mor:> ∀ `{SemiRing B},
        SemiRing_Morphism (naturals_to_semiring A B)
    ; naturals_initial:> Initial (bundle_semiring A) }.
\end{lstlisting}
Basing theory and programs on this abstract interface instead of on specific implementation (such as the ubiquitous Peano naturals \lstinline|nat| in the Coq standard library) is not only cleaner mathematically, but also facilitates easy swapping between implementations. And this benefit is far from theoretical, as diverse representations of the natural numbers are abound; for instance, unary, binary, factor multisets, and arrays of native machine words.

Since initial objects in categories are isomorphic, we can easily derive that \lstinline|naturals_to_semiring| gives isomorphisms between different \lstinline|Naturals| implementations:
\begin{lstlisting}
  Lemma iso_naturals `{Naturals A} `{Naturals B}:
    ∀ a: A, naturals_to_semiring B A (naturals_to_semiring A B a) = a.
\end{lstlisting}
This is very useful, because some properties of and operations on naturals are more easily proved, respectively defined, for concrete implementations (such as \lstinline|nat|) and then \emph{lifted} to the abstract \lstinline|Naturals| interface so that they work for arbitrary implementations. For example, while showing decidability for an arbitrary \lstinline|Naturals| implementation directly is tricky, it is very easy to show decidability for \lstinline|nat|. Using \lstinline|iso_naturals|, the latter can be very straightforwardly used to implement the former.
